Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2014-07-16 paper.tex
Some more details on the current state of the isolated BCs.

All of this can be done for the gravitational version of Poisson's equation. Even though we do not commonly speak of negative gravitational charges, they are simply a mathematical tool for obtaining the correct potential. In general,
\begin{equation}
  \sigma = \frac{1}{4\pi G} {\bm{\nabla}} \phi \cdot \hat{\mathbf{n}},
\end{equation}
where $\hat{\mathbf{n}}$ is the unit vector perpendicular to (and pointing away from) the surface. On the computational domain, using an appropriate discretized representation of the ${\bm{\nabla}}$ operator, we calculate $\sigma$ at every boundary location. This requires some care. In CASTRO, the boundaries are on cell edges a distance $dx / 2$ away from the outermost domain location. It is therefore tempting to use a central difference that uses the ghost zone on one side and the outermost zone inside the domain in the other. This is normally second-order accurate. However, by construction, the gradient of $\phi$ is discontinuous across that surface; it is zero just outside the surface, and non-zero just inside the surface. This discontinuity is what establishes the non-zero surface mass density, similar to how a jump in the electric field is associated with a surface charge density in electrostatics. Therefore it is inappropriate to use the central difference method. Instead, a one-sided difference is required, using only the data interior to the surface. For the discretization to be second-order accurate, data from the two nearest cells must be used.

 Therefore instead of using a difference centered on the boundary edge, we take a one-sided difference using the grid points just inside the domain. In order for the discretization to remain second-order accurate, data from the two closest grid points are used.}

We can then find the resulting potential on the boundary by summing over all of the surface charges:
\begin{equation}
  \Psi_{\text{l,m,n}} = \sum_{\text{l$^\prime$,m$^\prime$,n$^\prime$}} \frac{G\sigma_{\text{l$^\prime$,m$^\prime$,n$^\prime$}}}{\left|\mathbf{x}_{\text{l,m,n}} - \mathbf{x}_{\text{l$^\prime$,m$^\prime$,n$^\prime$}}\right|} dA_{\text{l$^\prime$,m$^\prime$,n$^\prime$}}
\end{equation}
Here (l, m, n) refer only to boundary locations. The self-contribution from a zone must be excluded in this sum, since the denominator would diverge there.

>>>>

All of this can be done for the gravitational version of Poisson's equation. Even though we do not commonly speak of negative gravitational charges, they are simply a mathematical tool for obtaining the correct potential. In general,
\begin{equation}
  \sigma = \frac{1}{4\pi G} {\bm{\nabla}} \phi \cdot \hat{\mathbf{n}},\label{eq-isolated-sigma}
\end{equation}
where $\hat{\mathbf{n}}$ is the unit vector perpendicular to (and pointing away from) the surface. On the computational domain, using an appropriate discretized representation of the ${\bm{\nabla}}$ operator, we calculate $\sigma$ at every boundary location. This requires some care. In CASTRO, the boundaries are on cell edges a distance $dx / 2$ away from the outermost domain location. It is therefore tempting to use a central difference that uses the ghost zone on one side and the outermost zone inside the domain in the other. This is normally second-order accurate. However, by construction, the gradient of $\phi$ is discontinuous across that surface; it is zero just outside the surface, and non-zero just inside the surface. This discontinuity is what establishes the non-zero surface mass density, similar to how a jump in the electric field is associated with a surface charge density in electrostatics. Therefore it is inappropriate to use the central difference method. Instead, a one-sided difference is required, using only the data interior to the surface. For the discretization to be second-order accurate, data from the two nearest cells must be used. Generally, $\left.\partial f/\partial x\right|_{\rm i} = (-3 f_{\rm i} + 4 f_{\rm i+1} - f_{\rm i+2}) / (2 \Delta x)$ is a second-order accurate forward difference to the derivative of a function $f$ at zone index i; however, we are indexing from the domain edge, which does not coincide with a cell center. We can use this formula on our finite volume grid if we let $\Delta x \to \Delta x / 2$ (here we assume that the index 0 corresponds to the left-most cell center):
\begin{equation*}
  \left.\frac{\partial \phi}{\partial x}\right|_{-1/2} = \frac{-3 \phi_{-1/2} + 4 \phi_{0} - \phi_{1/2}}{2(\Delta x / 2)} + \mathcal{O}(\Delta x^2).
\end{equation*}
To proceed, we note that $\phi_{-1/2} = 0$ by construction, and $\phi_{1/2} = (\phi_{1} + \phi_{0}) / 2$ to second order, so:
\begin{equation}
  \left.\frac{\partial \phi}{\partial x}\right|_{-1/2} = \frac{7\phi_{0} - \phi_{1}}{2\Delta x} + \mathcal{O}(\Delta x^2).
\end{equation}
A similar formula holds for the one-sided (backward) difference at the opposite boundary. Once this gradient has been calculated, we can calculate surface masses using Equation \ref{eq-isolated-sigma} (recalling to negate the result to obtain positive masses, following the above discussion). For example, for a mass on the lower surface of constant $x$,
\begin{equation}
  M_{-1/2} = -\sigma_{-1/2} \Delta A = \frac{1}{4\pi G} \frac{7\phi_{0} - \phi_{1}}{2\Delta x}\Delta y \Delta z.
\end{equation}
According to Gauss' law, the total mass induced on the surface should be equal to the mass enclosed inside the boundary. We have verified directly that the sum of the mass added on the surface approaches the total mass inside the domain at a second-order rate of convergence.

We can then find the resulting potential on the boundary by summing over all of the surface charges:
\begin{equation}
  \Psi_{\text{l,m,n}} = \sum_{\text{l$^\prime$,m$^\prime$,n$^\prime$}} g_{\text{l-l$^\prime$},\text{m-m$^\prime$},\text{n-n$^\prime$}} M_{\text{l$^\prime$,m$^\prime$,n$^\prime$}}
%= \sum_{\text{l$^\prime$,m$^\prime$,n$^\prime$}} \frac{GM_{\text{l$^\prime$,m$^\prime$,n$^\prime$}}}{\left|\mathbf{x}_{\text{l,m,n}} - \mathbf{x}_{\text{l$^\prime$,m$^\prime$,n$^\prime$}}\right|}
\end{equation}
Here (l, m, n) refer only to boundary locations, and $g_{\text{a},\text{b},\text{c}}$ is the Green's function for the Poisson equation. At large distances, this is simply $G / r$, where $G$ is the gravitational constant and $r$ is the distance. However, this analytical expression is not correct for the discretized Laplacian, and the corresponding Green's function is a series in the odd powers of $1/r$. We use the expression derived by \citet{burkhart:1997}, only retaining the terms up to $r^{-5}$. Also, this expression nominally diverges at $r = 0$, which physically corresponds to the self-contribution to the potential from a zone. However, this singularity is integrable: the cell does have a contribution to its own potential. A numerical evaluation of the integral of the Green's function over a cubic region yields $\phi(r = 0) = 2.38\, G M / \Delta x$, assuming $\Delta x = \Delta y = \Delta z$, and so our Green's function at the origin is $g_{0,0,0} = 2.38\, G / \Delta x$. This is similar to how other authors treat this singularity, for example \citet{hockney_eastwood}. However, they simply set the Green's function as $G / \Delta x$, that is with unit prefactor, without further explanation.
